---
title: "Forecasting 2020 US Election Outcomes: The Role of Employment, Education, gender, and Race"
subtitle: "My subtitle if needed"
author: 
  - Sean Liu
thanks: "Code and data are available at: https://github.com/Sluuu/Politics.git"
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(boot)
library(broom.mixed)
library(collapse)
library(gutenbergr)
library(janitor)
library(knitr)
library(marginaleffects)
library(modelsummary)
library(rstanarm)
library(tidybayes)
library(tidyverse)
```


# Introduction

we are interested in the US political support. In particular we are interested in whether we can forecast who a respondent is likely to vote for, based on knowing their employment status, highest level of education, and race. That means we are interested in a dataset with variables for who an individual voted for, and some of their characteristics, such as employment status, education level, gender, and their race. We use @citeR and @rohan.

The remainder of this paper is structured as follows. @sec-data....

# Features
The original CES 2020 dataset, which shows in @tbl-raw, contains 61000 observations and many variables. We chose to feature these 6 variables: "votereg", "CC20_410", "race", "educ", "employ", "gender", in our analysis. 

1. votereg: Shows whether the respondent is registered to vote or not.

2. CC20_410: Shows the president candidate the respondents voted for.

3. race: The racial or ethnic group best describe the respondents.

4. educ: the highest education level of the respondents have obtained.

5. employ: the current employment status of the respondent.

6. gender: the gender of the respondents 


# Data {#sec-data}

Our primary data is from the 2020 Cooperative Election Study, CES @CES2020. The data includes a nationally representative sample of 61,000 American adults. @CES2020 includes the data from the survey, a full guide to the data, and the questionnaires.

```{r}
#| label: fig-education 
#| fig-cap: filtered by gender 
#| echo: false
#| warning: false
#| message: false

ces2020 |>
  ggplot( aes(x = education, fill = voted_for)) +
  stat_count(position = "dodge") +
  facet_wrap(facets = vars(gender)) +
  theme_minimal() +
  labs(
    x = "Highest education",
    y = "Number of respondents",
    fill = "Voted for"
  ) +
  coord_flip() +
  scale_fill_brewer(palette = "Set1") +
  theme(legend.position = "bottom")

```


```{r}
#| label: fig-race 
#| fig-cap: filtered by gender 
#| echo: false
#| eval: false

  ggplot(ces2020, aes(x = race, fill = voted_for)) +
  stat_count(position = "dodge") +
  facet_wrap(facets = vars(gender)) +
  theme_minimal() +
  labs(
    x = "race",
    y = "Number of respondents",
    fill = "Voted for"
  ) +
  coord_flip() +
  scale_fill_brewer(palette = "Set1") +
  theme(legend.position = "bottom")
```
```{r}
#| label: fig-employment status
#| fig-cap: filtered by gender 
#| echo: false
#| eval: false
ces2020 |>
  ggplot(aes(x = employment_stat, fill = voted_for)) +
  stat_count(position = "dodge") +
  facet_wrap(facets = vars(gender)) +
  theme_minimal() +
  labs(
    x = "race",
    y = "Number of respondents",
    fill = "Voted for"
  ) +
  coord_flip() +
  scale_fill_brewer(palette = "Set1") +
  theme(legend.position = "bottom")
```

Talk more about it.



And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)




Talk way more about it. 



# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.  

\begin{align} 
y_i|\pi_i &\sim \mbox{Bern}(\pi_i) \\
\mbox{logit}(\pi_i) &= \alpha + \beta_1 \times \mbox{employemnt_status}_i + \beta_2 \times \mbox{race}_i + \beta_3 \times \mbox{education}_i + \beta_4 \times \mbox{gender}_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta_1 &\sim \mbox{Normal}(0, 2.5) \\
\beta_2 &\sim \mbox{Normal}(0, 2.5) \\
\beta_3 &\sim \mbox{Normal}(0, 2.5) \\
\beta_4 &\sim \mbox{Normal}(0, 2.5)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.


### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.


# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("starter_folder-main/models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "Supported Biden" = first_model
  ),
  statistic = "mad"
)
```

```{r}
#| echo: false
#| eval: true
#| label: fig-modelresultsfig
#| fig-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelplot(first_model, conf_level = 0.9) +
  labs(x = "90 per cent credibility interval")
```



# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this. 

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}


# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows... 

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

analysis_data <- arrow::read_parquet(here::here("starter_folder-main/data/analysis_data/analysis_data.parquet"))

set.seed(853)

analysis_data <- 
  analysis_data |> 
  slice_sample(n = 1000)


pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics

@fig-traceplot is a trace plot. It shows... This suggests...

@fig-rhatplot is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-traceplot
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot"]


plot(first_model, "trace")

```

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-rhatplot
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Rhat plot"]



plot(first_model, "rhat")
```



\newpage


# References


